Fractal Patterns and Music

### 2019-12-16 paper2.tex
Lazy update

\title{Fractal Patterns and Music}
\date{MATH-538, December 2019}
\author{Javier, Chris Benito, Nils Olsson}

@@

In this paper we explain the current state of fractals in music. We
include a description of fractals and how they can be perceived in music. We
discuss how the note frequencies follow a power law relation across many genres
of western music. We present the Michael Bulmer’s technique for creating music
that is close to true pink noise. There is also an overview of how fractals have
occurred in musical compositions in the past, including the specific example in
Bach’s Cello Suite No. 3. Then we created a program to perform dimension
calculations of music.

@@

A single key on a piano, when struck, produces a single pitch. Pitch is one of
the most fundamental properties of notes in music, but what exactly is pitch?
The sound that is produced by the piano is really the combination of several
strings vibrating, each producing different frequencies of sound. Together this
combination forms the characteristic sound of the piano (also called timbre, the
specific quality to the sound of, in this case, the piano). The strongest
frequency in this collection is called the fundamental frequency of the “note”
being played, and to this frequency a name is assigned. In the case of a piano,
and of all Western music, these frequency names come from the alphabet: A, B, C,
D, E, F, and G. Though pitches are unique unto their frequency, some pitches are
considered equivalent with respect to the ratio of their frequencies. For
example, the pitch A4 (concert A) sounds at 440 cycles per second (hertz), while
A5 sounds at 880 hertz: a ratio of 1:2. This ratio is called an octave, and thus
we have names for distances between frequencies which we call intervals. Within
Western music notation there are a finite number of named intervals; here’s a
few: perfect octaves, perfect fifths, diminished fifths, augmented fourths
(equivalent to diminished fifths), major thirds, minor thirds, etc. Between any
two notes (and even between a note and itself) there is an interval between
them, and is measured aptly by counting the number of keys of the piano it takes
to get from one note to the other where these steps are called “semitones”. For
example, the interval between any F and the nearest B on the piano consists of
six semitones. As an aside, this interval is known as a tritone and is one of
the most dissonant intervals producible on the piano. Returning to octaves,
despite having unique frequencies, any two pitches which are an interval of an
octave apart (or multiple octaves apart, e.g. C2 and C5) from one another, are
considered equivalent. Lastly, If you are familiar with the piano then you will
know there are actually are more than seven keys in between octaves, twelve
precisely. These are just smaller subdivisions of the frequency ranges between
octaves.

The other most fundamental property of notes is duration, also called rhythmic
value (or simply “value”). Music as listened to is inherently ephemeral: in the
moments in which we experience music by listening to it, it exists, but then it
is gone. How long we experience the sound of a pitch is called its duration, and
we might measure the duration of a pitch by how long it is played in seconds.
Establishing a baseline duration is important in the composition of music; some
pieces of music consist of pitches that are long in duration, while others
consist of pitches that exist only for a split-second. But not all pitches need
have the same duration, and so we have rhythm: the ratios between pitches in
terms of duration. For example we might play one pitch for twice as long as the
one before, and the next two pitches for a quarter as long. Like with pitch,
different rhythmic values also have names within Western music notation, some of
the most basic of which being whole notes, half notes, quarter notes ♩, and
eighth notes ♪. Notice that the names all have to do with fractions. A piece of
music might set a baseline speed by stating that the duration of a quarter note
should be a certain number of seconds—then the duration of a half note would be
twice as long as the duration of the quarter note, a whole note four times as
long, an eighth note half as long, a sixteenth note a quart as long, etc.
It is these two fundamental properties that are a part of every note in a piece of music.

@@

@@

The most dramatic limitation of this procedure and this metric of scaling in
general is that it requires the musical material to be monophonic---only one
note at a time---while the vast majority of modern music (and actually the
majority of music since the 14th century) is polyphonic. Despite this, a piece
of music containing simultaneous notes across simultaneous voices could
potentially be reconstructed into standalone parts, where each roughly functions
as standalone piece consisting of only a monophonic melody. Think of a Bach
cantata for soprano, alto, tenor and bass four-part choir, where each voice is
essentially its own piece, its own melody. On the other hand, not all music can
be so simply reconstructed into parts. Even the Bach selection examined by Su
and Wu (the Bourrée), despite being almost entirely monophonic, contains
polyphony in measures 2, 4, and 28 (the last), but this may be an artifact of
the editor, as some copies have a single G instead of the E chord. Either way,
one can simplify a polyphonic melody by selecting a single constituent note
whenever there's a simultaneous group of notes, as I did for simplifying the
Bourrée.

@@

>>>>

\title{Fractal Patterns and Music}
\date{MATH-538, December 2019}
\author{Javier Anguiano, Chris Benito, Nils Olsson}

@@

In this paper we will explore the history and some existing research in the area
of fractals in music. We include a description of fractals and how they can be
perceived in music, as well as a primer on pitch, rhythm, melody and harmony
(the basic building-blocks of music). We discuss how the note frequencies follow
a power law relation across many genres of western music. We present the Michael
Bulmer’s technique for creating music that is close to true pink noise. There is
also an overview of how fractals have occurred in musical compositions in the
past, including the specific example in Bach’s Cello Suite No. 3. Then we
attempted to reproduce the results of a paper in calculating a particular
fractal dimension within a musical example.

\begin{center}
    \hrulefill{}
\end{center}

@@

A single key on a piano, when struck, produces a single pitch. Pitch is one of
the most fundamental properties of notes in music, but what exactly is pitch?
The sound that is produced by the piano is really the combination of several
strings vibrating, each producing different frequencies of sound. Together this
combination forms the characteristic sound of the piano (also called timbre, the
specific quality to the sound of, in this case, the piano). The strongest
frequency in this collection is called the fundamental frequency of the “note”
being played, and to this frequency a name is assigned. In the case of a piano,
and of all Western music, these frequency names come from the alphabet: A, B, C,
D, E, F, and G.

Though pitches are unique unto their frequency, some pitches are
considered equivalent with respect to the ratio of their frequencies. For
example, the pitch A4 (concert A) sounds at 440 cycles per second (hertz), while
A5 sounds at 880 hertz: a ratio of 1:2. This ratio is called an octave, and thus
we have names for distances between frequencies which we call intervals. Within
Western music notation there are a finite number of named intervals; here are a
few: perfect octaves, perfect fifths, diminished fifths, augmented fourths
(equivalent to diminished fifths), major thirds, minor thirds, etc. Between any
two notes (and even between a note and itself) there is an interval between
them, and is measured aptly by counting the number of keys of the piano it takes
to get from one note to the other where these steps are called “semitones”. For
example, the interval between any F and the nearest B on the piano consists of
six semitones. As an aside, this interval is known as a tritone and is one of
the most dissonant intervals producible on the piano.

Returning to octaves,
despite having unique frequencies, any two pitches which are an interval of an
octave apart (or multiple octaves apart, e.g. C2 and C5) from one another, are
considered equivalent. Lastly, If you are familiar with the piano then you will
know there are actually are more than seven keys in between octaves, twelve
precisely. These are just smaller subdivisions of the frequency ranges between
octaves.

The other most fundamental property of notes is duration, also called rhythmic
value (or simply “value”). Music as listened to is inherently ephemeral: in the
moments in which we experience music by listening to it, it exists, but then it
is gone. How long we experience the sound of a pitch is called its duration, and
we might measure the duration of a pitch by how long it is played in seconds.
Establishing a baseline duration is important in the composition of music; some
pieces of music consist of pitches that are long in duration, while others
consist of pitches that exist only for a split-second.

However, not all pitches need
have the same duration, and so we have rhythm: the ratios between pitches in
terms of duration. For example we might play one pitch for twice as long as the
one before, and the next two pitches for a quarter as long. Like with pitch,
different rhythmic values also have names within Western music notation, some of
the most basic of which being whole notes, half notes, quarter notes, and
eighth notes. Notice that the names all have to do with fractions. A piece of
music might set a baseline speed by stating that the duration of a quarter note
should be a certain number of seconds—then the duration of a half note would be
twice as long as the duration of the quarter note, a whole note four times as
long, an eighth note half as long, a sixteenth note a quart as long, etc.

In summary, it is these two fundamental properties, pitch and rhythmic value,
that are a part of every note in a piece of music.

@@

(All the material of this project can be found at
\url{https://github.com/nilsso/math538-project}, including all source code and
the paper itself.)

The Python language was chosen for the attempt at reproducing the results in the
Su and Wu paper, and the first stage was to develop a way to represent a melody
in code. There is a plethora of music notation software out there in the world,
but the especially niche \href{http://lilypond.org/}{LilyPond} system allows one
to notate music in plain text. Due to familiarity with this system, LilyPond
notation was adopted.

First a few helper classes, \verb+Note+ and \verb+Melody+, were written to
encapsulate the data about pitch and rhythmic values for notes and to
encapsulate data about the notes in a melody, as well as facilitate the parsing
of LilyPond notation (abbreviated Ly). In the end, constructing a melody can be
accomplished like in the following example:

\begin{minted}{python}
    notes = [ "c'4", "g", "e8", "c" ]
    M = Melody(notes)

    for n in M.notes:
        print(n)
\end{minted}

Which results in the formatted output:

\begin{minted}{text}
    C5 quarter-note
    G4 quarter-note
    E4 eighth-note
    C4 eighth-note
\end{minted}

Then chosen from the three examples in the article was Gossec's \emph{Gavotte},
which the melody was simplified and notated using the Ly format:

\begin{minted}{py}
notes = [
        # 1
        "a'8", "b", "a", "fis", "g", "a", "g", "e"
        # 2
        "d4", "b'", "d,2",
        # 3
        "g8", "a", "g", "e", "fis", "g", "fis", "d",
        # etc...
\end{minted}

Finally functions for calculating the point distribution $p_i(r)$, the local Hölder
exponent $\alpha$, and the Hölder variation $f(\alpha)$ were developed, as well
as plotting utilities. The end result of these steps are
Figures~\ref{fig:results_1},~\ref{fig:results_2} and~\ref{fig:results_3}
that begin to
approach those illustrated in the Su and Wu article.

\begin{figure}[ht!]
    \centering
    \includegraphics[width=\linewidth]{figures/results01.png}
    \caption{%
        Melodic interval sequence of the Gossec's \emph{Gavotte}.
    }\label{fig:results_1}
\end{figure}
\begin{figure}[ht!]
    \centering
    \includegraphics[width=\linewidth]{figures/results02.png}
    \caption{%
        Scaling of $p_i(r)$ with sub-cover size $r$ of \emph{Gavotte}.
    }\label{fig:results_2}
\end{figure}
\begin{figure}[ht!]
    \centering
    \includegraphics[width=\linewidth]{figures/results03.png}
    \caption{%
        Variation of Hölder exponents along the melody sequence converted from
        the \emph{Gavotte}.
    }\label{fig:results_3}
\end{figure}

@@

The most dramatic limitation of this procedure and this metric of scaling in
general is that it requires the musical material to be monophonic---only one
note at a time---while the vast majority of modern music (and actually the
majority of music since the 14th century) is polyphonic. Despite this, a piece
of music containing simultaneous notes across simultaneous voices could
potentially be reconstructed into standalone parts, where each roughly functions
as standalone piece consisting of only a monophonic melody. Think of a Bach
cantata for soprano, alto, tenor and bass four-part choir, where each voice is
essentially its own piece, its own melody. On the other hand, not all music can
be so simply reconstructed into parts. Even the Bach selection examined by Su
and Wu (the Bourrée), despite being almost entirely monophonic, contains
polyphony in measures 2, 4, and 28 (the last), but this may be an artifact of
the editor, as some copies have a single G instead of the E chord. Either way,
one can simplify a polyphonic melody by selecting a single constituent note
whenever there's a simultaneous group of notes, as was done for simplifying the
Bourrée.

@@

In conclusion, there is quite a lot of insight into the form and structure of
music that can be gained by studying music with an appreciation of fractals.
Fractal geometry patterns have a way of appearing in most of nature, so despite
(most) music being intelligently designed it should not be all too surprising
that fractals also show up in music, especially when there exist psychoacoustic
explanations as to why humans have been composing music in certain genres and
with certain sensibilities for melodic and harmonic consonance and dissonance
for centuries.
The composition of music is also accessible in a very mathematically intuitive
and intriguing way by applying stochastic and self-scaling patterns.
Lastly, there is much room for further research and improvement of self-scaling
pattern analysis techniques, with applications in classifying individual pieces
or even entire genres of music based on their different fractal dimensions.
